Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2023-10-16 manu-draft.qmd
working on results from RN comments

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.

A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]) and @eq-ebase as a forward calculating model with estimated values for $a$, $R$, and $b$. All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b) were also estimated with the synthetic time series. The time series was estimated using the observed data described above, the separate metabolic components and their parameters in @eq-ebase, and approximate metabolic rates from the Odum open-water method as implemented in the WtRegDO R package [@Beck21] as an approximate estimate for back-calculation of the EBASE parameters. In short, $P$, $R$, and $D$ as $\text{mmol}/\text{m}^2/\text{d}$ were estimated using the Odum method and the separate synthetic components for EBASE were created by estimating the corresponding parameters, such that $P /PAR = a$, $R$ as is, and $D = k_w\left(C_{sat} - C \right)$, where $k_w$ includes the $b$ parameter. Daily estimates for each were created as the daily average for each parameter fit with a polynomial spline to create a smoothed output. The synthetic dissolved oxygen time series was then created as the integrated sum of the separate components for $P$, $R$, and $D$ divided by water column depth. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

>>>>

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series of dissolved oxygen concentration created from observed water temperature, salinity, air temperature, $PAR$, and wind speed measured every 15 minutes. These assessments evaluated if 1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year (2021) of continuous data from Apalachicola Bay, Florida. Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point). Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual data versus time to create a complete dataset from January 1st to December 31st.

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimatd by 1) applying the Odum open-water method as implemented in the WtRegDO R package \[@Beck21\], 2) computing monthly averages of the resulting daily estimates of $P$ and $R$, 3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and 4) by fitting a spline function to the monthly averages of $R$ and $a$.

The resulting synthetic time series of light efficiency, dissolved oxygen budget terms ($P$, $D$, and $R$), and dissolved oxygen concentration are shown in @fig-synapa. The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a). The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to that of light efficiency, but noisier, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air–water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity. The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c). There is a prominent diurnal cycle that is superimposed on subseasonal variability, presumably driven by radiation and winds.

#### Recovery of model parameters

#### Sensitivity to priors
